On Extending a Full-Sharing Multithreaded Tabling Design with Batched Scheduling

### 2015-04-16 paper.tex
removing stuff
abstract - ok

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage[noend]{algorithmic}
\usepackage{algorithm}
\usepackage{epsfig}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{subfigure}
\usepackage{url}
\newcommand{\verbatimproperties}{\renewcommand{\baselinestretch}{0.85} \small}
%\newcommand{\tableproperties}{\centering \big}
\newcommand{\LINECOMMENT}[1]{$\triangleright$ #1}

%\setlength{\belowcaptionskip}{-26pt}

@@

\title{On Extending a Full-Sharing Design with Batched Scheduling}

@@

\begin{abstract}

  \textbf{Keywords:} Multithreading, Tabling, Concurrency, Batched
  Scheduling
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Tabling~\cite{Chen-96} is an implementation technique that overcomes
some limitations of traditional Prolog systems in dealing with
redundant sub-computations and recursion. Tabling consists of storing
intermediate answers for subgoals so that they can be reused when a
repeated subgoal appears during the resolution process. Tabling has
become a popular and successful technique thanks to the
ground-breaking work in the XSB Prolog system and in particular in the
SLG-WAM engine~\cite{Sagonas-98}, the most successful engine of
XSB. The success of SLG-WAM led to several alternative implementations
that differ in the execution rule, in the data-structures used to
implement tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

Multithreading in Prolog is the ability to concurrently perform
computations, in which each computation runs independently but shares
the program clauses. When multithreading is combined with tabling, we
have the best of both worlds, since we can exploit the combination of
higher procedural control with higher declarative semantics. To the
best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
were the only Prolog systems that were able to combine both
multithreading with tabling. Yap implements a SWI-Prolog compatible
multi-threading library~\cite{Wielemaker-03}. Yap's threads have their
own execution stacks and only share the code area where predicates,
records, flags and other global non-backtrackable data are stored. For
tabled evaluation, a thread views its tables as private but, at the
engine level, Yap has three designs that vary from a \emph{No-Sharing}
(NS) design, where each thread allocates fully private tables for each
new tabled subgoal called during its computation, to a
\emph{Full-Sharing} (FS), where threads share the complete table
space.

In this paper we discuss a novel approach for combining the FS design
with batched scheduling and we present a performance analysis
comparison between local scheduling with batched
scheduling. Experimental results show that despite the extra data
structures required to combine FS design with batched scheduling, the
execution time of combination is still quite competitive when
comparing against the FS with local scheduling.

The remainder of the paper is organized as follows. First, we briefly
introduce some background and related work. Then, we describe our
approach for combining batched scheduling with the FS design. Next, we
show the most important implementation details and finally, we discuss
experimental results and we end by outlining some conclusions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Table Space Organization}

@@

\begin{wrapfigure}{r}{4.5cm}
\vspace{-\intextsep}
\centering
\includegraphics[width=4.5cm]{figures/table-space.pdf}
\caption{Table space organization}
\label{fig_table_space}
\vspace{-0.5\intextsep}
\end{wrapfigure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Yap's Multithreaded Tabling Support}

Yap implements a SWI-Prolog compatible multi-threading
library~\cite{Wielemaker-03}. Like in SWI-Prolog, Yap's threads have
their own execution stacks and only share the code area where
predicates, records, flags and other global non-backtrackable data are
stored.

In a multithreaded tabling evaluation, a thread views its tables as
private but, at the engine level, Yap has three designs that vary from
a \emph{No-Sharing} (NS) design, where each thread allocates fully
private tables for each new tabled subgoal called during its
computation, to a \emph{Full-Sharing} (FS), where threads share the
complete table space. Figure~\ref{fig_table_space} shows Yap's
multithreaded tabling support with an abstraction layer to which
threads interact (the threads represented in the bottom of the
figure), and at the engine level the configuration of the NS and FS
designs in a situation where several different threads evaluated the
same tabled subgoal call $i$.

In the NS design, each thread allocates fully private tables for each
new subgoal called during its computation. One can observe that the
table entry data structure still stores the common information for the
predicate (such as the arity or the evaluation strategy), and then
each thread $t$ has its own cell $T_t$ inside a \emph{bucket array}
which points to the private data structures.

\begin{figure}[!ht]
\vspace{-\intextsep}
\centering
\includegraphics[width=9.5cm]{figures/yap-mt.pdf}
\caption{Yap's multithreaded tabled space using the NS and FS designs}
\label{fig_yap_mt_support}
\vspace{-1.0\intextsep}
\end{figure}

In the FS design, part of the subgoal frame information (the
\emph{subgoal entry} data structure) and the answer trie structure are
now shared among all threads. The previous subgoal frame data
structure was split into two: the \emph{subgoal entry} stores common
information for the subgoal call (such as the pointer to the shared
answer trie structure); the remaining information (the \emph{subgoal
  frame} data structure) remains private to each thread.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scheduling Strategies}

The decision about the evaluation flow is determined by the
\emph{scheduling strategy}. Different strategies may have a
significant impact on performance, and may lead to a different
ordering of solutions to the query goal. Arguably, the two most
successful tabling scheduling strategies are \emph{local scheduling}
and \emph{batched scheduling}~\cite{Freire-96}.

Local scheduling strategy schedules the evaluation of a program in a
breath-first manner. It favors the backtracking first with completion
instead of the forward execution, leaving the consumption of answers
for last. Thus, it only allows a Cluster of Dependent Subgoals (CDS)
to return answers only after the completion point has been
reached~\cite{Freire-96}. %% In other words, the local scheduling tries
%% to keep a CDS as minimal as possible. When new answers are found, they
%% are added to the table space and the computation fails as consequence,
%% tabled subgoals inside a CDS propagate their answers to outside the
%% CDS only after its completion point is found.
Local scheduling causes a sooner completion of subgoals, which creates
less complex dependencies between them.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extending Full-Sharing with Batched Scheduling}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Our Approach}

@@

\begin{figure}[!ht]
\vspace{-\intextsep}
\centering
\includegraphics[width=10.5cm]{figures/pcc.pdf}
\caption{The \emph{FS} design with the PAC procedure - (a) private
  chaining and (b) public chaining}
\label{fig_tabtries_pcc}
\vspace{-1.0\intextsep}
\end{figure}

@@

\setlength{\tabcolsep}{12pt}
\begin{table}[!ht]
\vspace{-\intextsep}
\centering
%% \caption{Overhead ratios, when compared with the NS design with $1$
%%   thread (running with local scheduling without TabMalloc) for the NS
%%   and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
%%   threads with local and batched scheduling on the five sets of
%%   benchmarks (best ratios by row and by design for the Minimum,
%%   Average and Maximum are in bold)}
\caption{Overhead ratios, when compared with the NS design with 1
  thread for the NS and FS designs, when running 1, 8, 16, 24 and 32
  threads with local and batched scheduling (lowest is the best and it
  is marked in bold by row and by design for the Minimum, Average and
  Maximum)}
\vspace{-0.5\intextsep}
%\scalebox{1.00} {
\begin{tabular}{ll|cc|cc}
\hline\hline
\multicolumn{2}{c|}{\multirow{2}{*}{\bf Threads}} &
\multicolumn{2}{c|}{\multirow{1}{*}{\bf NS}} &
\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS}}\\ \cline{3-6}
& 
& \multicolumn{1}{c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}
& \multicolumn{1}{|c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}\\
\hline
\multirow{4}{*}{\bf 1}
& {\bf Min }& {\bf 0.53}& 0.55& 1.01& {\bf 0.95}\\
& {\bf Avg }& {\bf 0.78}& 0.82& {\bf 1.30}& 1.46\\
& {\bf Max }& 1.06& {\bf 1.05}& {\bf 1.76}& 2.33\\
& {\bf StD }& 0.15& 0.14& 0.22& 0.44\\
\hline
\multirow{4}{*}{\bf 8}
& {\bf Min }& 0.66& {\bf 0.63}& 1.16&{\bf  0.99}\\
& {\bf Avg }& {\bf 0.85}& 0.88& {\bf 1.88}& 1.95\\
& {\bf Max }& {\bf 1.12}& 1.14& {\bf 2.82}& 3.49\\
& {\bf StD }& 0.13& 0.14& 0.60& 0.79\\
\hline
\multirow{4}{*}{\bf 16}
& {\bf Min }& 0.85& {\bf 0.75}& 1.17& {\bf 1.06}\\
& {\bf Avg }& {\bf 0.98}& 1.00& {\bf 1.97}& 2.08\\
& {\bf Max }& {\bf 1.16}& 1.31& {\bf 3.14}& 3.69\\
& {\bf StD }& 0.09& 0.17& 0.65& 0.83\\
\hline
\multirow{4}{*}{\bf 24}
& {\bf Min }& {\bf 0.91}& 0.93& 1.16& {\bf 1.09}\\
& {\bf Avg }& {\bf 1.15}& 1.16& {\bf 2.06}& 2.19\\
& {\bf Max }& 1.72& {\bf 1.60}& {\bf 3.49}& 4.08\\
& {\bf StD }& 0.20& 0.21& 0.70& 0.91\\
\hline
\multirow{4}{*}{\bf 32}
& {\bf Min }& 1.05& {\bf 1.04}& 1.33& {\bf 1.26}\\
& {\bf Avg }& 1.51& {\bf 1.49}& {\bf 2.24}& 2.41\\
& {\bf Max }& {\bf 2.52}& 2.63& {\bf 3.71}& 4.51\\
& {\bf StD }& 0.45& 0.45& 0.74& 1.02\\
\hline\hline
\end{tabular}%}
\label{tab_batched_overhead}
\vspace{-0.7\intextsep}
\end{table}

>>>>

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage[noend]{algorithmic}
\usepackage{algorithm}
\usepackage{epsfig}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{subfigure}
\usepackage{url}
\newcommand{\verbatimproperties}{\renewcommand{\baselinestretch}{0.85} \small}
%\newcommand{\tableproperties}{\centering \big}
\newcommand{\LINECOMMENT}[1]{$\triangleright$ #1}
%\setlength{\belowcaptionskip}{-26pt}

@@

\title{On Extending a Full-Sharing Multithreaded Tabling Design with
  Batched Scheduling}

@@

\begin{abstract}
Tabling is an implementation technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Multithreading in Prolog is the
ability to concurrently perform computations, in which each
computation runs independently but shares the program clauses. When
multithreading is combined with tabling, we have the best of both
worlds, since we can exploit the combination of higher procedural
control with higher declarative semantics.  To support this
combination, the Yap system has at engine level, multiple designs that
vary from a No-Sharing design, where each thread allocates fully
private tables for each new tabled subgoal called during its
computation, to a Full-Sharing design, where threads share the
complete table space. Arguably, batched scheduling is one of the most
successful tabling scheduling strategies. In this work, we propose an
extension of the Full-Sharing design to support batched
scheduling. Experimental results show that, despite the extra data
structures required to support the extension, the FS design remains
quite competitive.

\textbf{Keywords:} Logic Programming, Multithreading, Tabling, Scheduling.
\end{abstract}

\section{Introduction}

Tabling~\cite{Chen-96} is an implementation technique that overcomes
some limitations of traditional Prolog systems in dealing with
redundant sub-computations and recursion. Tabling consists of storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of XSB. The
success of SLG-WAM led to several alternative implementations that
differ in the execution rule, in the data-structures used to implement
tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

Multithreading in Prolog is the ability to concurrently perform
computations, in which each computation runs independently but shares
the program clauses. When multithreading is combined with tabling, we
have the best of both worlds, since we can exploit the combination of
higher procedural control with higher declarative semantics. To the
best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
are the only Prolog systems that support the combination of
multithreading with tabling. Yap implements a SWI-Prolog compatible
multithreading library~\cite{Wielemaker-03}. Yap's threads have their
own execution stacks and only share the code area where predicates,
records, flags and other global non-backtrackable data are stored. For
tabled evaluation, a thread views its tables as private but, at the
engine level, Yap has three designs~\cite{Areias-12a} that vary from a
\emph{No-Sharing} (NS) design, where each thread allocates fully
private tables for each new tabled subgoal called during its
computation, to a \emph{Full-Sharing} (FS) design, where threads share
the complete table space.

The decision about the evaluation flow is determined by the
\emph{scheduling strategy}. Different strategies may have a
significant impact on performance, and may lead to a different
ordering of solutions to the query goal. Arguably, the two most
successful tabling scheduling strategies are \emph{local scheduling}
and \emph{batched scheduling}~\cite{Freire-96}. Local scheduling tries
to complete subgoals as soon as possible. When new answers are found,
they are added to the table space and the evaluation fails. Answers
are only returned when all program clauses for the subgoal at hand
were resolved. Batched scheduling favors forward execution first,
backtracking next, and consuming answers or completion last. It thus
tries to delay the need to move around the search tree by batching the
return of answers to repeated subgoals. When new answers are found for
a particular tabled subgoal, they are added to the table space and the
evaluation continues.

With the FS design, all tables are shared. Thus, since several threads
can be inserting answers in the same table, when an answer already
exists, it is not possible to determine if the answer is new or
repeated for a certain thread without further support. However, for
local scheduling this is not a problem because, for repeated and new
answers, local scheduling always fails. The problem is with batched
scheduling that requires further support to work with the FS design
since, with batched scheduling, only repeated answers should fail. We
have thus to detect when a new found answer is also a new answer for
the thread at hand in order to ensure that new answers for a given
subgoal call are correctly propagated one and only once to all
thread's calls.

In this work, we propose an extension to the table space data
structures, which we named \emph{Private Answer Chaining (PAC)}, as a
way to keep track, per thread and subgoal call, of the answers that
were already found and propagated to all repeated calls. We discuss in
detail our proposal for extending the FS design with batched
scheduling and we present a performance analysis comparison between
local and batched scheduling. Experimental results show that, despite
the extra PAC data structures required to support batched scheduling
with the FS design, the execution time of the combination is still
quite competitive when comparing against the FS design with local
scheduling.

The remainder of the paper is organized as follows. First, we briefly
introduce some background and related work. Then, we describe our
approach for combining batched scheduling with the FS design and we
discuss the most important implementation details. Finally, we present
experimental results and we end by outlining some conclusions.

\section{Background}

@@

\subsection{Table Space Organization}

@@

\begin{wrapfigure}{r}{4.5cm}
%\vspace{-\intextsep}
\centering
\includegraphics[width=4.5cm]{figures/table-space.pdf}
\caption{Table space organization}
\label{fig_table_space}
%\vspace{-0.5\intextsep}
\end{wrapfigure}

@@

\subsection{Yap's Multithreaded Tabling Support}

In Yap, a thread views its tables as private but, at the engine level,
it uses a common table space where tables are shared among all
threads. Yap implements three designs for concurrent tabling support
that vary from a \emph{No-Sharing} (NS) design, where each thread
allocates fully private tables for each new tabled subgoal called
during its computation, to a \emph{Full-Sharing} (FS) design, where
threads share the complete table space. Figure~\ref{fig_table_space}
shows Yap's multithreaded table space organization for the NS and FS
designs, where an interface layer abstracts the design being used at
the engine level. The figure illustrates the main differences between
the two designs for a situation where several threads are evaluating
the same tabled subgoal call $call\_i$.

\begin{figure}[!ht]
%\vspace{-0.5\intextsep}
\centering
\includegraphics[width=9.5cm]{figures/yap-mt.pdf}
\caption{Yap's multithreaded table space organization for the NS and FS designs}
\label{fig_yap_mt_support}
%\vspace{-1.0\intextsep}
\end{figure}

When using the NS design, each thread allocates fully private tables
for each new subgoal called during its computation. One can observe
that the table entry data structure still stores the common
information for the predicate (such as the arity or the evaluation
strategy), and then each thread $t$ has its own cell $T_t$ inside a
\emph{bucket array} which points to the private data structures.

When using the FS design, the subgoal and answer trie structures and
part of the subgoal frame information (the \emph{subgoal entry} data
structure in Fig.~\ref{fig_yap_mt_support}) are shared among all
threads. The previous subgoal frame data structure was split in two:
the \emph{subgoal entry} stores common information for the subgoal
call (such as the pointer to the shared answer trie structure); the
remaining information is kept private to each thread in the
\emph{subgoal frame} data structure.

\subsection{Scheduling Strategies}

Local scheduling schedules the evaluation of a program in a
breath-first manner. It favors backtracking first with completion
instead of forward execution, leaving the consumption of answers for
last. Local scheduling only allows a \emph{Cluster of Dependent
  Subgoals} (CDS) to return answers after a fixpoint point has been
reached~\cite{Freire-96}.

In other words, local scheduling tries
to keep CDS as minimal as possible. When new answers are found, they
are added to the table space and the computation fails as consequence,
tabled subgoals inside a CDS propagate their answers to outside the
CDS only after its completion point is found.

Local scheduling causes a sooner completion of subgoals, which creates
less complex dependencies between them.

@@

\section{Extending Full-Sharing with Batched Scheduling}

@@

\subsection{Our Approach}

@@

\begin{figure}[!ht]
%\vspace{-\intextsep}
\centering
\includegraphics[width=10.5cm]{figures/pcc.pdf}
\caption{The PAC procedure using (a) private chaining and (b) public
  chaining}
\label{fig_tabtries_pcc}
%\vspace{-1.0\intextsep}
\end{figure}

@@

\setlength{\tabcolsep}{12pt}
\begin{table}[!ht]
%\vspace{-\intextsep}
\centering
%% \caption{Overhead ratios, when compared with the NS design with $1$
%%   thread (running with local scheduling without TabMalloc) for the NS
%%   and FS designs (with TabMalloc), when running 1, 8, 16, 24 and 32
%%   threads with local and batched scheduling on the five sets of
%%   benchmarks (best ratios by row and by design for the Minimum,
%%   Average and Maximum are in bold)}
\caption{Overhead ratios, when compared with the NS design with 1
  thread for the NS and FS designs, when running 1, 8, 16, 24 and 32
  threads with local and batched scheduling (lowest is the best and it
  is marked in bold by row and by design for the Minimum, Average and
  Maximum)}
%\vspace{-0.5\intextsep}
%\scalebox{1.00} {
\begin{tabular}{ll|cc|cc}
\hline\hline
\multicolumn{2}{c|}{\multirow{2}{*}{\bf Threads}} &
\multicolumn{2}{c|}{\multirow{1}{*}{\bf NS}} &
\multicolumn{2}{|c}{\multirow{1}{*}{\bf FS}}\\ \cline{3-6}
& 
& \multicolumn{1}{c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}
& \multicolumn{1}{|c}{\bf Local}
& \multicolumn{1}{c}{\bf Batched}\\
\hline
\multirow{4}{*}{\bf 1}
& {\bf Min }& {\bf 0.53}& 0.55& 1.01& {\bf 0.95}\\
& {\bf Avg }& {\bf 0.78}& 0.82& {\bf 1.30}& 1.46\\
& {\bf Max }& 1.06& {\bf 1.05}& {\bf 1.76}& 2.33\\
& {\bf StD }& 0.15& 0.14& 0.22& 0.44\\
\hline
\multirow{4}{*}{\bf 8}
& {\bf Min }& 0.66& {\bf 0.63}& 1.16&{\bf  0.99}\\
& {\bf Avg }& {\bf 0.85}& 0.88& {\bf 1.88}& 1.95\\
& {\bf Max }& {\bf 1.12}& 1.14& {\bf 2.82}& 3.49\\
& {\bf StD }& 0.13& 0.14& 0.60& 0.79\\
\hline
\multirow{4}{*}{\bf 16}
& {\bf Min }& 0.85& {\bf 0.75}& 1.17& {\bf 1.06}\\
& {\bf Avg }& {\bf 0.98}& 1.00& {\bf 1.97}& 2.08\\
& {\bf Max }& {\bf 1.16}& 1.31& {\bf 3.14}& 3.69\\
& {\bf StD }& 0.09& 0.17& 0.65& 0.83\\
\hline
\multirow{4}{*}{\bf 24}
& {\bf Min }& {\bf 0.91}& 0.93& 1.16& {\bf 1.09}\\
& {\bf Avg }& {\bf 1.15}& 1.16& {\bf 2.06}& 2.19\\
& {\bf Max }& 1.72& {\bf 1.60}& {\bf 3.49}& 4.08\\
& {\bf StD }& 0.20& 0.21& 0.70& 0.91\\
\hline
\multirow{4}{*}{\bf 32}
& {\bf Min }& 1.05& {\bf 1.04}& 1.33& {\bf 1.26}\\
& {\bf Avg }& 1.51& {\bf 1.49}& {\bf 2.24}& 2.41\\
& {\bf Max }& {\bf 2.52}& 2.63& {\bf 3.71}& 4.51\\
& {\bf StD }& 0.45& 0.45& 0.74& 1.02\\
\hline\hline
\end{tabular}%}
\label{tab_batched_overhead}
%\vspace{-0.7\intextsep}
\end{table}
